Icicle: write once, run once

### 2016-05-25 2016fhpc/section/02-Source.tex
Hide syntax examples to save space

Let-bindings allow a name to be used in place of the expression, in the rest of some query.
\begin{code}
let diff = open - close
in  mean diff
\end{code}

The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.
Count can be expressed by the following fold:
\begin{code}
fold count = 0 then count + 1
\end{code}

Filters take a predicate, which can be thought of as a stream of booleans, and a query to perform for the satisfying values.
The following query will count the number of entries where the open price exceeds the close price.
\begin{code}
filter open > close in count
\end{code}

Filters correspond to @WHERE@ clauses in SQL, except that the predicate does not apply to the entire top-level query, only to the subquery.
This makes it slightly easier to define queries like the proportion of entities satisfying some predicate to all entities: 
\begin{code}
(filter open > close in count) / count
\end{code}

Group takes a key to group by, and a query to perform on the values of each grouping.
This query groups the entries into buckets of the difference between open and close, and counts the number of records in each bucket.
\begin{code}
group ((open - close) / 100) in count
\end{code}

$\mi{Def}$ is used for function and query definitions.
The @count@ function takes no arguments and returns a fold counting the number of elements.
\begin{code}
function count
 = fold c = 0 then c + 1;
\end{code}

Sum can be defined taking a single argument, @value@, which is the elements to compute the sum of.
It returns a fold that starts at zero and for every element, increases the old sum by @value@.
\begin{code}
function sum (value : Element Int)
 = fold s = 0 then s + value;
\end{code}

The definition of mean is just the sum divided by the count.
\begin{code}
function mean (value : Element Int)
 = sum value / count;
\end{code}

Queries must return a single value, and are given it a name:
\begin{code}
query count_all_records = count;
\end{code}

@@

\subsection{Type system}

The typing rules for Icicle are given in figure~\ref{fig:source:type:exp}.
Judgment $\Typecheck{\Gamma}{\mi{Exp}}{\TauMode}$ means that under context $\Gamma$, the expression has a given value type and modality.
Judgment $\TypecheckP{\mi{Prim}}{\TauFun}$ associates a primitive with its function type.
Judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
Judgment $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and returns an environment containing the function or query name and its type.
Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and returns a context containing only the types of the queries.

>>>>

% Let-bindings allow a name to be used in place of the expression, in the rest of some query.
% \begin{code}
% let diff = open - close
% in  mean diff
% \end{code}

The fold form takes the name of the fold binding, and two expressions: the initial value and the update value for each element.

% Count can be expressed by the following fold:
% \begin{code}
% fold count = 0 then count + 1
% \end{code}

Filters take a predicate, which can be thought of as a stream of booleans, and a query to perform for the satisfying values.

% The following query will count the number of entries where the open price exceeds the close price.
% \begin{code}
% filter open > close in count
% \end{code}

% Filters correspond to @WHERE@ clauses in SQL, except that the predicate does not apply to the entire top-level query, only to the subquery.
% This makes it slightly easier to define queries like the proportion of entities satisfying some predicate to all entities: 
% \begin{code}
% (filter open > close in count) / count
% \end{code}

Group takes a key to group by, and a query to perform on the values of each grouping.

% This query groups the entries into buckets of the difference between open and close, and counts the number of records in each bucket.
% \begin{code}
% group ((open - close) / 100) in count
% \end{code}

$\mi{Def}$ is used for function and query definitions.
% The @count@ function takes no arguments and returns a fold counting the number of elements.
% \begin{code}
% function count
%  = fold c = 0 then c + 1;
% \end{code}

% Sum can be defined taking a single argument, @value@, which is the elements to compute the sum of.
% It returns a fold that starts at zero and for every element, increases the old sum by @value@.
% \begin{code}
% function sum (value : Element Int)
%  = fold s = 0 then s + value;
% \end{code}

% The definition of mean is just the sum divided by the count.
% \begin{code}
% function mean (value : Element Int)
%  = sum value / count;
% \end{code}

% Queries must return a single value, and are given it a name:
% \begin{code}
% query count_all_records = count;
% \end{code}

@@

% ---------------------------------------------------------
\subsection{Type system}
The typing rules for Icicle are given in figure~\ref{fig:source:type:exp}.
Judgment $\Typecheck{\Gamma}{\mi{Exp}}{\TauMode}$ means that under context $\Gamma$, the expression has a given value type and modality.
Judgment $\TypecheckP{\mi{Prim}}{\TauFun}$ associates a primitive with its function type.
Judgment $\TypecheckApp{\TauFun}{\ov{\TauMode}}{\TauMode}$ is used for automatically boxing and unboxing modalities in function application: a function type applied to a list of argument value types and modalities results in a single value type and modality.
Judgment $\TypecheckS{\Gamma}{\mi{Def}}{\Gamma}$ takes an input environment and function or query, and returns an environment containing the function or query name and its type.
Finally, $\TypecheckS{}{\mi{Top}}{\Gamma}$ takes a top-level definition with a table, functions and queries, and returns a context containing only the types of the queries.
